```{r CH3-chunkopts, include=FALSE, echo = FALSE}

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, background = "#707070", fig.align = "center", cache = TRUE)

```

```{r CH3-setup, include=FALSE}

## This chapter .Rmd code relies on two sourced R codes: 
## 'bird_data_processing.R' and 'bird_data_analysis.R.' These do the
## heavy lifting before importing results here for ggplot-ifying and presenting
## in Results and Discussion.

## Data manipulation --> analysis --> Rmarkdown (this doc)


## Load sourced R codes
all.current <- TRUE

if (all.current == TRUE) {

  load(file = "bird_data_analysis_complete_02272020.RData")

}

if (all.current == FALSE) {
  # This takes a little while, maybe 15 min. depending on computer
  source("bird_data_analysis.R")
  
}


## Load required libraries
  library(ggplot2)
  library(ggeffects)
  library(rgl)
  library(pander)
  library(reshape2)
  library(gridExtra)
  library(ggrepel)
  library(tibble)
  library(vegan)
  library(stringr)
  library(kableExtra)
  library(knitr)
  library(dplyr)
  library(sjPlot)


  options(digits = 3)

  
# For outputting in latex; use \setstretch{1.05} for headings and \setstretch{1.65} for the main text. Also use: \chaptermark{Passerine community}; \widowpenalty10000; \clubpenalty10000; \raggedbottom

```


# Conserving native trees increases native bird diversity and (shifts) community composition on office developments

## Abstract---reduce to 250 words

In cities, vegetation provides critical shelter, nesting, and foraging habitat for bird species of interest. Human actions---including development and landscaping choices---determine vegetation community composition and structure. 
Thus, human actions impacting vegetation are critically important to bird conservation in cities. A better understanding of how bird communities are impacted by parcel-scale actions that alter vegetation communities can help guide policy and management best practices to improve matrix habitat quality and quantity.
Here, I examined how bird habitat use varies along a vegetation gradient created by different development and landscaping choices. Where previous research focused on public green space and private residential property, I expand this research to 20 commercial office developments in King County, Washington, USA selected using stratified random sampling. 
I quantified the bird communities and observed feeding behavior on office developments, collected data on tree, shrub, and ground cover communities, and gathered information about neighborhood scale socio-economic and land cover variables. 
I found that all measures of the bird community are positively influenced by the presence of more native conifers, particularly old trees that are protected during development activities, and the presence of native shrubs. Measures of the bird community are negatively influenced by higher non-native tree density.
In contrast to prior research, top-down neighborhood-scale variables did not explain variation in any measures of the bird community on office developments. Importantly, I found that birds are associated with the same habitat on office developments as observed elsewhere. Together, my findings suggest an important role for developers, land owners, landscape architects, and tree protection policy in bird conservation.

## Introduction

Vegetation is a critical component of bird habitat; vegetation composition and structure are long-recognized predictors of species presence and diversity [@cody1981; @macarthur1961; @betts2013; @jones2012; @paker2014]. In cities, vegetation provides critical shelter, nesting, and foraging habitat for bird species of interest [@antikainen1992; @goldstein1986; @narango2017].

Human actions in urban areas including development and landscaping choices determine vegetation community composition and structure [@faeth2011; @grimm2017; @niinemets2008; @dyson2019veg]. Development actions are disturbances and include removing native vegetation and topsoil and compacting soil with heavy equipment. Landscaping actions partly replace succession processes and include choosing and planting tree and shrub species in specified proportions and locations [@grimm2017; @niinemets2008]. The resulting vegetation communities may differ in the proportion of native species and ornamental non-native species as well as tree canopy cover in historically forested areas [@dyson2019veg].

Thus, the habitat available for birds in cities is almost entirely driven by human choices [@lepczyk2004; @muller2013; @vanheezik2012; @hamer2008; @nudds1977]. Habitat use is determined by species' habitat selection process and adaptability, the spatial layout of habitat in the landscape, habitat availability and quality, and the local species pool [@cody1985; @muller2013]. Sites with more native vegetation community composition are more likely to retain native bird species, as many native birds prefer to feed in native trees and shrubs [@paker2014; @chace2006; @belaire2014; @lepczyk2004; @chong2014; @pennington2011], though some are more adaptable and will readily use non-native ornamental vegetation [@belaire2014; @daniels2006]. Heterogenous plant layers are positively correlated with high bird diversity and native forest birds [@beissinger1982; @paker2014; @daniels2006; @chong2014; @smith2015; @cooke2002; @kroll2014; @cooke2002; @evans2009; @melles2003; @pennington2011]. Where vegetation structure is simplified, bird abundance is reduced, particularly for leaf-gleaning insectivores [@betts2013; @ellis2012]. Large lawns and monoculture woods without underbrush are favorable for alien bird species and urban exploiters [@paker2014; @smith2015]. One potential contributing factor to this pattern is that ground cover and increased vegetation structural complexity are frequently associated with increased abundance and diversity of arthropods [@frampton2000]. 

Habitat loss and degradation are primary drivers of extinction in anthropogenic habitat [@alberti2005; @polasky2005; @rosenzweig2003]. Successful conservation likely requires the urban habitat (vegetation) matrix to provide enough resources for local species in addition to habitat preservation [@miller2002; @fahrig2001; @fischer2006; @polasky2005; @rosenzweig2003; @shoffner2018]. Acting to alter urban vegetation communities can positively impact bird species richness and assemblages [@threlfall2016]. A better understanding of how bird communities are impacted by human actions at the parcel-scale that alter vegetation communities can help guide policy and management best practices to improve matrix habitat quality and quantity. [@goddard2010; @miller2002]. Native birds are of conservation interest to ecologists & concerned citizens [@belcher2019]. In the Puget Sound region of Washington, these include species of concern including Red-breasted Sapsucker, Pileated Woodpecker and Evening Grosbeak, as well as charismatic birds associated with native conifers including Chestnut-backed Chickadee, Golden-crowned Kinglet, Brown Creeper, and Varied Thrush.

Here, I examined how bird habitat use varies along a vegetation gradient created by different development and landscaping choices [@dyson2019veg]. Where previous research has focused on understanding this relationship for public green space [e.g. @antikainen1992; @threlfall2016] or private residential property [e.g. @daniels2006; @narango2017; @goddard2010], I expand this research to commercial office developments.

I hypothesized that the species richness and community composition of birds found on office developments, as well as the community of foraging birds, would be explained by variables describing the outcome human actions impacting vegetation at the parcel-scale [@belaire2014; @galitsky2015; @daniels2006]. Based on previous research, I also hypothesized that bird communities would be explained by top-down neighborhood-scale socio-economic and land cover variables found significant in studies of residential land use types, including median household income [@lerman2011; @luck2013; @melles2003; @schutz2015; @beumer2015; @vanheezik2013]. Previous research suggests that for birds in the Puget Trough local factors were twice as important as landscape patterns for frugivores, seed eaters, ground foragers, and bark foragers [@galitsky2015]. Therefore, I also hypothesized that parcel-scale variables would be more important in explaining variation in bird communities than neighborhood-scale variables. A local understanding of the relative influence these factors in determining habitat use by birds is an important consideration for urban bird conservation planning and success [@kinzig2005]. 

To address these questions, I quantified the bird communities and observed feeding behavior on office developments, collected data on tree, shrub, and ground cover communities, and gathered information about neighborhood scale socio-economic and land cover variables.
I surveyed winter bird communities as they are less well studied than breeding bird communities [@jokimaki2003] and resources obtained during the winter season are critical for survival and impact body condition for migration and reproduction [@faaborg2010; @manuwal1987; @devries2008; @marra1998]. 

I found that all measures of the bird community are positively influenced by the presence of more native conifers, particularly old trees that are protected during development activities, and the presence of native shrubs. Measures of the bird community are negatively influenced by higher non-native tree density. In contrast to prior research, top-down neighborhood-scale variables did not explain variation in any measures of the bird community on office developments. Importantly, I found that birds are associated with the same habitat on office developments as observed elsewhere. Together, my findings suggest an important role for developers, land owners, landscape architects, and tree protection policy in bird conservation, and a need for additional research examining how land use policy is implemented at the parcel-scale, what factors influence developer and landowner actions, and how to best incentivize actions like native conifer preservation associated with more diverse winter bird communities.


## Methods

### Study area

Bellevue (2017 population 144,000) and Redmond (population 64,000) are located east of Seattle in King County, Washington in the Puget Trough region [@Census2017]. This temperate lowlands region lies between the Cascades and Puget Sound, with post glaciation conifer-dominated forests of Douglas-fir, western red cedar, and western hemlock (_Pseudotsuga menziesii_, _Thuja plicata_, and _Tsuga heterophylla_). Both cities share a similar ecological history including glaciation ending 10,000 years ago, a similar disturbance timeline for logging and agriculture, and have grown considerably since the opening of the Evergreen Point Floating Bridge (SR 520) in 1963. They are at similar elevations (< 500 ft) and experience the same climate and weather. The two cities are approximately 20 miles from the foothills of the Cascades, and 30 miles from forested mountains > 4,000 ft elevation. Two large lakes, Lake Washington and Lake Sammamish, are adjacent to city borders (Figure 1).

```{r map}

# Create map here.

```

To sample the entire vegetation gradient, I used a disproportionate stratified random sampling approach [more detail see @dyson2019veg]. To avoid confounding factors, I restricted the sampling pool to sites in the 25^th^ to 85^th^ percentile for size and the 15^th^ to 85^th^ percentile for surrounding impervious surfaces. I also limited the sampling frame to exclude high-rise buildings in Bellevue's central business district and properties south of I-90.

I requested property access by sending three rounds of mailings to the property owner or manager on file [@dyson2019private; @dyson2019veg]. Twenty of the 46 requests sent were accepted, 6 requests were rejected, and 20 received no response or were not deliverable. My sample included 3-5 sites from each of the five vegetation strata, with `r length(which(all.covariates$DistrictName == "BELLEVUE")) ` sites in Bellevue and `r length(which(all.covariates$DistrictName == "REDMOND")) ` of the study sites in Redmond. The study sites ranged from `r round(min(all.covariates$acres), 3)` acres to `r round(max(all.covariates$acres), 3)` acres. The office developments on study sites were built between 1975 and 2008, with some fully leased to tenants and other partly or fully owner-occupied. Commercial use of the sites included light industrial, office space, and medical/dental offices. Company size ranged from one employee to tens of thousands.

### Bird surveys

I sampled birds 4 times at each of the 20 sites for two seasons; winter season sampling occured between December 12^th^ 2014 and March 1^st^ 2015 and between December 11^th^ 2015 and February 27^th^ 2016. Site visit order was randomized for each of the 4 rounds of surveys, except when scheduling conflicts arose with the property owner or manager. While these scheduling constraints may introduce bias, they are an unavoidable when working on private property [@dyson2019private]. Sampling occurred on days when wind speed was less than 10 mph and rainfall was less than 0.1''. Repeated sampling helped reduce winter weather bias; while I delayed sampling when there was thick fog, light rain and fog are normal winter weather conditions for the region [@marzluff2016causal]. 

I used the standardized search method, which allows for comparison of heterogeneous study sites, retains the parcel as the unit of analysis instead of changing it to the point count circle, and is compatible with the geometry of parcels [@dyson2019veg; @watson2003]. The method also encourages observer movement, which allowed for detection of cryptic winter species, observation of bird behavior and foraging activity, and repositioning to mitigate the impacts of urban noise. Pilot data collected using both point count and standardized search methodology confirmed the superiority of the standardized search methodology in detecting winter birds (Supplemental Table 1). 

I visited two sites per day. Sweeps at the first site began at sunrise (approx. 8:00 AM) and sweeps at the second site finished between 10:00AM and 1:00PM due to the variable sampling length and between site transit times. During each site visit, 20-minute sweeps were conducted consecutively until no new species observed in the sweep. This resulted in `r min(bird.by.visit.PA.info$number.visit.sweeps) `--`r max(bird.by.visit.PA.info$number.visit.sweeps) ` sweeps per visit (median `r median(bird.by.visit.PA.info$number.visit.sweeps) ` sweeps per visit). I recorded each bird species seen and whether birds were observed foraging. Foraging activities required visual confirmation except for woodpeckers. I did not record birds flying overhead, water dependent birds, or raptors [@melles2003; @jokimaki1998]. I recorded sweeps using an Olympus WS-822 digital recorder; myself and Bob Sundstrom confirmed identifications made in the field and performed quality checks.


### Bird guilds

I constructed bird guilds based on diet, foraging substrate, and forest preference following previous research [@bnao]. Observed bird species included insectivores [insects 90% or greater proportion of diet; @blair2008; @zhou2012; @beissinger1982; @chace2006], grainivores [seeds 90% or greater; @blair2008; @werner2009; @zhou2012; @chace2006] and omnivores [mix of invertebrates, fruit, seeds, and plant parts; @jokimaki2003; @lepczyk2004]. Foraging substrates included ground foragers and tree and shrub foragers [@beissinger1982; @chace2006]. Forest preference included conifer, mixed conifer/deciduous, and no preference or open habitat preferred. 


### Independent variables

**Detection variables:** I recorded loud noise levels and weather conditions that could have impacted my ability to detect birds [including overcast skies, fog, and drizzle; @leukering1998]. I used median hourly data for air temperature, wind speed, and solar radiation (brightness), and cumulative rainfall from AgWeatherNet [Woodinville sensor; @agweathernet]. 

**Vegetation:** Individual trees > 4" DBH were identified to species or genus, and all members of the shrub community were identified to species or genus [@dyson2019veg; @sunset1995; @krussmann1985; @symonds1963shrub; @symonds1958tree; @dirr2009manual; @dirr1997hardy; @cope2001muenscher; @sibley2009; @city-trees]. Tree and shrub genera were assigned to one of three provenance categories---native, non-native, or ambiguous---based on range information from the United States Geological Survey and the United States Department of Agriculture [@USGSplants; @USDAplants]. The ambiguous category was used for genera including both native and non-native cultivated species that are difficult to distinguish, and/or frequently interbred and sold as crosses. 

Native and ornamental tree community types and shrub community types were identified with flexible beta clustering [@dyson2019veg]. The tallest five native conifers were measured and the median was used as a proxy for tree age [@dyson2019private]. The presence of stands predating development was determined from historical records [@dyson2019veg]. Other variables derived from the shrub community include the density of native shrubs and effective species richness of native shrub species [@jost2006].

**Ground cover:** Ground cover types were recorded on paper maps, then hand digitized to calculate the area of impervious surfaces on site [@QGIS].

**Parcel and Neighborhood Variables:** I derived parcel-scale and neighborhood-scale variables from available databases  [Table 1; @ACS5yr; @taxDB; @KCDB; @xian2011; @homer2015; @dyson2019veg]. I chose a 500 m buffer around the parcel boundary for neighborhood-scale variables based on previous research and the likely winter foraging ranges of local native bird species [@bnao; @melles2003].

I quantified the percent of tall tree canopy cover and of short and medium height vegetation cover using 2015 NAIP, 2014 Redmond LiDAR data and 2016 King County LiDAR data [@NAIP; @RedLiDAR; @KingCoLiDAR]. I classified pixels as tall tree canopy when NDVI $\geq$ 0.15 and object height $\geq$ 50 feet and as short and medium height vegetation when NDVI $\geq$ 0.15 and object height < 50 feet. I calculated the number of major intersections by intersecting major road data, checking for accuracy by hand, and then counting the number of intersections within a 500 m radius [@roadsdata].

**Maintenance data:** I interviewed each site's landscaping service using the elite interview technique and supplemented these interviews with observational data from site visits [@dexter1970; @harvey2011]. I also censused each site for logs (over 6" diameter), snags (over 6" diameter and over 6' tall), and stumps (tree stumps over 6" diameter). Stumps where the tree had grown from a co-dominant trunk were counted. 



```{r CH3-covariate-definitions}

covariate.definitions <- tibble(
  `Variable Name` = c(
  "1.",
  "Loud Noise (% visits)",
  "Fog (% visits)",
  "Light Rain (% visits)",
  "Overcast (% visits)",
  "Median Air Temperature (F)",
  "Median Wind Speed (MPH)",
  "Median Wind Gust (MPH)",
  "Total Precipitation (inches)",
  "Median Solar Radiation (W/m^2)",
  
  "2.",
  "Area",
  "Building Age (in 2017)",
  "Appraised Land Value per Acre",
  
  
  "3.",
  "Impervious w/in 500 m (%)",
  "Tall Vegetation w/in 500 m (%)",
  "Short and Medium Vegetation w/in 500 m (%)",
  "Major intersections w/in 500 m (count)",
  "Median Income",
  "Percent Foreign-Born",
  
  
  "4.",
  "Impervious on Site (%)",
  "Stands Predate Development",
  "Median Height of Dominant Conifers",
  
  "Irrigation",
  "Dead Wood",
  
  "Tree Community Group",
  "Shrub Community Group",
    
  "Tree Density",
  "Non-native Tree Effective Species Richness",
  "Non-native Tree Density",
  "Native Tree Effective Species Richness",
  "Native Tree Density",
  "Native Conifer Density",
  
  "Shrub Density",
  "Non-native Shrub Effective Species Richness",
  "Non-native Shrub Density",
  "Native Shrub Effective Species Richness",
  "Native Shrub Density"

  
  ),
  
  Definition = c(
  "DETECTION VARIABLES",
  "Site visits with loud noise.",
  "Site visits with fog.",
  "Site visits with light rain.",
  "Site visits where cloud cover exceeded 75%.",
  "Median of median air temperature observed during each visit.",
  "Median of median wind speed observed during each visit.",
  "Median of median wind gusts observed during each visit.",
  "Precipitation summed over all 8 visits.",
  "Median of median solar radiation observed during each visit.",
  
  
  "SITE-SCALE SOCIO-ECONOMIC VARIABLES",
  "Site area, in acres.",
  "Age of building on site (or mean age for multiple buildings) in 2017.",
  "Appraised land value divided by site area. One missing assessed land value was replaced with the population median land value.",
  
  
  "NEIGHBORHOOD LAND COVER AND SOCIO-ECONOMIC VARIABLES",
  "Percent of impervious surface within 500 m of the site's perimeter.",
  "Percent of tall tree vegetation canopy cover within 500 m of the site's perimeter.",
  "Percent of short and medium vegetation canopy cover within 500 m of the site's perimeter.",
  "Number of major intersections located within 500 m of the site's perimeter.",
  
  "The median income of residents for the site's block group.",
  "The percent of residents born outside of the United States for the site's block group.",
  
  "DEVELOPMENT & LANDSCAPING CHOICE VARIABLES",
  "Percent of the parcel's area occupied by impervious surface.",
  "Presence/absence of a stand of three+ trees nearby one another that predate development. (Development)",
  "Median height (m) of 5 dominant native conifers; a proxy for age.",
  "Indicates whether irrigation is used during the summer months.",
  "Total abundance of stumps, logs, and snags on site.",
  
  "Tree community type (native or ornamental) based on flexible beta cluster analysis of tree density data.",
  "Shrub community type (native or ornamental) based on flexible beta cluster analysis of shrub density data.",
  
  "Total number of all trees on site divided by site area.",
  "Effective species richness of non-native trees. ESR is calculated as exp(H'), following Jost (2006).",
  "Total number of non-native trees on site divided by site area.",
  "ESR of native trees as exp(H')",
  "Total number of native trees on site divided by site area",
  "Total number of Douglas-fir, western red cedar, and western hemlock on site divided by site area.",
  
  "Total number of all shurbs on site divided by site area.",
  "ESR of non-native shrubs calculated as exp(H').",
  "Total number of non-native shrubs on site divided by site area.",
  "ESR of native shrubs calculated as exp(H').",
  "Total number of native shrubs on site divided by site area."

  
  ),
  
    `Data Source` = c(
"",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"WSU AgWeatherNet",
"WSU AgWeatherNet",
"WSU AgWeatherNet",
"WSU AgWeatherNet",
"WSU AgWeatherNet",

"",
"King County Assessor",
"King County Assessor",
"King County Assessor",

"",
"National Land Cover Database 2011 Percent Developed Imperviousness dataset updated in 2014",
"NDVI calculated using the 2015 NAIP data and object height calculated using 2014 and 2013 LiDAR data",
"NDVI calculated using the 2015 NAIP data and object height calculated using 2014 and 2016 LiDAR data",
"OpenStreetMap",
"American Community Survey 2014 5-year block group",
"American Community Survey 2014 5-year block group",

"",
"Site survey",
"Site survey",
"Site survey",
"Interviews and site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey",
"Site survey"
)
)

  

knitr::kable(covariate.definitions,
       booktabs = TRUE,
       longtable = TRUE,
       caption = "Definition of independent variables used in bird community analysis.") %>%
    kable_styling(full_width = F,
           latex_options = c("striped"),
           font_size = 10) %>%
    column_spec(1, bold = T, width = "12em")%>%
    column_spec(2, width = "20em")%>%
    column_spec(3, width = "12em")%>%
    row_spec(c(1,11,14,20,27), bold = T)

```




### Data analysis

I calculated bird incidence as $incidence = \frac{\text{count of visits observed}}{8 \text{ visits}}$. I used incidence and not a repeated measures design as between site, not between year, variability is important. Unknown species were recorded in the field but removed from analysis; unknown species accounted for only `r sum(bird.by.visit.PA.info$sweeps.present[which(bird.by.visit.PA.info$text.spp == "unknown")])` out of `r sum(bird.by.visit.PA.info$sweeps.present) ` observations (`r round(sum(bird.by.visit.PA.info$sweeps.present[which(bird.by.visit.PA.info$text.spp == "unknown")]) / sum(bird.by.visit.PA.info$sweeps.present) * 100, 3)`%).

Rare species observed fewer than two times were removed prior to calculating univariate effective species richness for all birds and bird guilds [@jost2006]. Rare species were not removed from multivariate community composition matrices [@mccune2002]. TITAN requires more than 3 observations of a species [@titan2]. No further transformations were used in this analysis.

I used PERMANOVA with the Euclidian and Bray-Curtis distance matrix respectively for univariate analysis of effective species richness and multivariate analysis of community composition,  [adonis2 {vegan}; @anderson2001; @bakker2012; @jost2006; @vegan; @mccune2002]. I first tested each independent variable in a simple model. Significance for all tests was assessed on pseudo-_F_ _p_-values at $\alpha \leq 0.05$. I used ANOVA to test for significant differences in categorical group dispersion using anova(betadisper()) {stats} {vegan} [@vegan]. Betadispersion was calculated for all categorical variables and evaluated at$\alpha \leq 0.05$.

For both univariate and multivariate analyses, I then constructed models using combinations of 1 to 4 of the significant variables identified. I used a custom AICc function for PERMANOVA to compare models and identify variables with the best support based on summed AIC weights [@dysoncode; @arnold2010]. In addition for univariate analysis, I examined the relationship between bird guild effective species richness and the significant variables identified. This approach helped ensure interpretable results that are more relevant to urban planners and management professionals. 

To explore changes in community composition along significant environmental gradients, I used point-biserial correlation [categorical variables; @de2013; @decaceres2010; @dysoncode] and TITAN analysis [@king2014; @baker2013; @titan2]. As point-biserial correlation is sensitive to initial conditions, I used a custom wrapper for the multipatt {indicspecies} implementation to run the function 100 times and keep indicator species present in > 50% of the runs and with mean statistic > 0.5 [@dysoncode; @hinkle2003]. 





## Results

I observed `r ncol(matrify.incidence.clean)` bird species, with `r sum(descriptive.bird.bysp$foraging.num.sites > 0) ` species observed foraging (Supplemental Tables XXX & XXX). The number of species observed on each site varied between `r min(descriptive.bird.bysite$bird.sp.richness)` and `r max(descriptive.bird.bysite$bird.sp.richness)` species (mean `r mean(descriptive.bird.bysite$bird.sp.richness)` $\pm$ `r round(sd(descriptive.bird.bysite$bird.sp.richness),1)`), with the number of bird species observed foraging between `r min(descriptive.bird.bysite$foraging.sp.richness)` to `r max(descriptive.bird.bysite$foraging.sp.richness)` species per site (mean `r mean(descriptive.bird.bysite$foraging.sp.richness)` $\pm$ `r round(sd(descriptive.bird.bysite$foraging.sp.richness),1)`). 

The most frequently observed species included Dark-eyed Junco (`r round(16.625/20*100,1)`% of visits), American Crow (`r round(15.125/20*100,1)`% of visits), and Golden-crowned Kinglet (`r round(14.500/20*100,1)`% of visits). Common Raven, Downy Woodpecker, Hutton's Vireo, Killdeer, and Rock Pigeon were each seen only once. Most bird species I observed were native; of the three non-native species encountered, only European Starling was observed at multiple sites (11 sites), while House Sparrow and Rock Pigeon were seen at only 1 site each.


### Effective species richness results

Variables describing site-scale impervious surface coverage, tree, and shrub communities significantly explained variation in univariate analysis of effective species richness (ESR) after controlling for the proportion of overcast survey visits, the only significant detection variable. Specifically, 
median dominant conifer height (m, pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "height.m.median","pseudo.F"]`),
the presence of stands predating development (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "stand.predate.development","pseudo.F"]`),
non-native tree density (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "tree.orn.dens","pseudo.F"]`),
tree community type (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "tree.cluster.name","pseudo.F"]`),
native conifer density (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "conifer.nat.dens","pseudo.F"]`),
native shrub effective species richness (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "shrub.nat.esr","pseudo.F"]`),
native shrub density (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "shrub.nat.dens","pseudo.F"]`),
impervious surface cover on site (%,  (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "impervious.pct","pseudo.F"]`)),
native tree density (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "tree.nat.dens","pseudo.F"]`), and
shrub community type  (pseudo-_F_ = `r sig.uPERM.results[sig.uPERM.results$var.names == "shrub.cluster.name","pseudo.F"]`)
were significant. 
No neighborhood-scale socio-economic or land cover variables were significant (Supplemental Table XXX). Significant variables suggest ecologically meaningful patterns; for example, increasing median dominant conifer height (m) by 10m increases expected bird ESR by 4 species, the presence of a tree stand predating development increases expected bird ESR by about 4.5 species, and increasing the non-native tree density by 10 trees/acre decreases expected bird ESR by about 2 species (Figure \@ref(fig:univariate-significant-plots)). Within the tree community, measures of native tree density positively impact predicted bird ESR, while non-native tree density negatively impacts it (Table \@ref(tab:tree-density-effect)). 



```{r univariate-significant-plots, fig.width = 7.5, fig.cap="(ref:univaritesig)"}

# Median conifer height
fit <-
  lm(data = univariate.sprich,
     bird.effective.sp_2 ~ clouds.76.100.proportion + height.m.median)

p.height.m.median <-
  plot_model(fit,
             type = "pred",
             terms = "height.m.median",
             show.data = TRUE,
             axis.title = c("Median Dominant Conifer Height (m)",
                            "Bird Effective Species Richness"),
             title = "")
        
# stand predates development
univariate.sprich$stand.predate.development.p <- 
  ifelse(univariate.sprich$stand.predate.development == "no.unkn",
                                              "No.or.Unknown",
                                              "Yes")
fit <-
  lm(data = univariate.sprich,
     bird.effective.sp_2 ~ clouds.76.100.proportion + stand.predate.development.p)

p.SPD <-
  plot_model(fit,
             type = "pred",
             terms = "stand.predate.development.p",
             show.data = TRUE, 
             axis.title = c("Stand Predates Development",
                            " "),
             title = "",) + scale_x_discrete(limits = c("No or Unknown","Yes"))

        
# ornamental tree density
fit <-
  lm(data = univariate.sprich,
     bird.effective.sp_2 ~ clouds.76.100.proportion + tree.orn.dens)

p.tree.orn.dens <-
  plot_model(fit,
             type = "pred",
             terms = "tree.orn.dens",
             show.data = TRUE,
             axis.title = c("Non-native Tree Density",
                            " "),
             title = "")
        
grid.arrange(p.height.m.median, p.SPD, p.tree.orn.dens, ncol = 3)


```


``` {r tree-density-effect}

        # ornamental tree density
        fit.tree.orn.dens <-  lm(data = univariate.sprich, bird.effective.sp_2 ~ clouds.76.100.proportion + tree.orn.dens)
        
        # native tree density
        fit.tree.nat.dens <-  lm(data = univariate.sprich, bird.effective.sp_2 ~ clouds.76.100.proportion + tree.nat.dens)
        
        # native CONIFER density
        fit.conifer.nat.dens <-  lm(data = univariate.sprich, bird.effective.sp_2 ~ clouds.76.100.proportion + conifer.nat.dens)
        
cols <- c("All Non-Native"="#E69F00","All Native"="#56B4E9","Native Conifer"="#0072B2")
        ggplot() + geom_ribbon(
          data = ggpredict(fit.tree.orn.dens, terms = "tree.orn.dens"),
          aes(x, ymin = conf.low, ymax = conf.high),
          fill = "#bbbbbb",
          alpha = .3
        ) +
          geom_ribbon(
            data = ggpredict(fit.tree.nat.dens, terms = "tree.nat.dens"),
            aes(x, ymin = conf.low, ymax = conf.high),
            fill = "#bbbbbb",
            alpha = .3
          ) +
          geom_ribbon(
            data = ggpredict(fit.conifer.nat.dens, terms = "conifer.nat.dens"),
            aes(x, ymin = conf.low, ymax = conf.high),
            fill = "#bbbbbb",
            alpha = .3
          ) +
          
          geom_line(
            data = ggpredict(fit.tree.orn.dens, terms = "tree.orn.dens"),
            aes(x, predicted, col = "All Non-Native"),
            size = 1.6,
          ) +
          geom_line(
            data = ggpredict(fit.tree.nat.dens, terms = "tree.nat.dens"),
            aes(x, predicted, col = "All Native"),
            size = 1.6
          ) +
          geom_line(
            data = ggpredict(fit.conifer.nat.dens, terms = "conifer.nat.dens"),
            aes(x, predicted, col = "Native Conifer"),
            size = 1.6
          ) +
          scale_colour_manual(name="Tree Type",values=cols) +
          ylab("Predicted Bird Effective Species Richness") +
          xlab("Site Tree Density (trees/acre)") +
          labs(title = "Effect of Tree Density on Predicted Bird ESR")
        

        
        
```

The variable with the most support is median dominant conifer height (Table \@ref(tab:uPERM-summed-AIC-weight)). The expanded model with the presence of a stand predating development recieves equal support using $\Delta$AICc, however summed AIC weights across all models tested suggests it may be an uninformative addition possibly because it is too similar [@arnold2010].

```{r uPERM-summed-AIC-weight}

multi.uPERM.AICc.weights.byvar$`Significant Variable` <- c("Native Conifer Density",
                                                           "Median Dom. Conifer Height (m)",
                                                           "Impervious on Site (%)",
                                                           "Native Shrub Density",
                                                           "Native Shrub ESR",
                                                           "Stand Predates Development",
                                                           "Native Tree Density",
                                                           "Non-native Tree Density",
                                                           "Shrub Community Type",
                                                           "Tree Community Type")

knitr::kable(multi.uPERM.AICc.weights.byvar, digits = 3)

```




#### Effective species richness of bird guilds {-}

Effective species richness of omnivores and insectivores were both positively correlated with presence of stands predating development and median Douglas-fir height (m) and negatively correlated with percent impervious on site, while grainivores were not correlated to any of these variables. Insectivores showed particularly high variation explained for stands predating development and median Douglas-fir height (Table \@ref(tab:CH3-bird-guild-esr-table)). Effective species richness of birds foraging on the ground and in trees or shrubs was positively correlated to stands predating development and median Douglas-fir height (m) though surprisingly not with percent impervious on site (adjusted _p_-value = 0.06). Only the effective species richness of conifer-associated birds was positively correlated to stands predating development and median Douglas-fir height and negatively to percent impervious on site; birds associated with mixed forests were not correlated with any variables and birds associated with open areas or without preference were correlated only weakly with median Douglas-fir height.

```{r CH3-bird-guild-esr-table, warning=FALSE}

pretty.bird.guild.esr <- tibble(
  `Bird Guild` = c("DIET",
           "  Omnivore",
           "  Insectivore",
           "  Grainivore",
           "FORAGING LOCATION",
           "  Ground",
           "  Tree and Shrub",
           "FOREST PREFERENCE",
           "  Conifer",
           "  Mixed",
           "  Open or No Preference"
           ),
  `Stand Predates Devlp't` = c("",
           paste(
                   #"Correllation = NA",
                   #round(guild.spd.results["omnivore.esr", 1], 2),
                   "Variation Expld: ",
                   round(guild.spd.results["omnivore.esr",
                               "var.explnd"], 3),
                   "Adj pval: ",
                   round(guild.spd.results["omnivore.esr",
                               "adj.pval"], 3)),
           paste(
                   #"Estimate = NA",
                   #round(guild.spd.results["insectivore.esr", 1], 2),
                   "Variation Expld: ",
                   round(guild.spd.results["insectivore.esr",
                               "var.explnd"], 3),
                   "Adj pval: ",
                   round(guild.spd.results["insectivore.esr",
                               "adj.pval"], 4)),
           paste0("Adj pval: ",
              round(guild.spd.results["grainivore.esr", "adj.pval"], 3)),
           "",
           paste(
                   #"Estimate = ",
                   #round(guild.spd.results["ground.for.esr", 1], 2),
                   "Variation Expld: ",
                   round(guild.spd.results["ground.for.esr",
                               "var.explnd"], 3),
                   "Adj pval: ",
                   round(guild.spd.results["ground.for.esr",
                               "adj.pval"], 4)),
           paste(
                   #"Estimate = ",
                   #round(guild.spd.results["treshrb.for.esr", 1], 2),
                   "Variation Expld: ",
                   round(guild.spd.results["treshrb.for.esr", 
                               "var.explnd"], 3),
                   "Adj pval: ",
                   round(guild.spd.results["treshrb.for.esr",
                               "adj.pval"], 4)),
           "",
           paste(
                   #"Estimate = ",
                   #round(guild.spd.results["conifer.esr", 1], 2),
                   "Variation Expld: ",
                   round(guild.spd.results["conifer.esr", 
                               "var.explnd"], 3),
                   "Adj pval: ",
                   round(guild.spd.results["conifer.esr",
                               "adj.pval"], 4)),
           paste0("Adj pval: ",
              round(guild.spd.results["mixed.esr", "adj.pval"], 3)),
           paste0("Adj pval: ",
              round(guild.spd.results["open.esr", "adj.pval"], 3))
           ),
  
  `Median D-f Height (m)` = c("",
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["omnivore.esr", 
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["omnivore.esr",
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["omnivore.esr",
                                 "adj.pval"], 4)
                   ), 
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["insectivore.esr",
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["insectivore.esr", 
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["insectivore.esr",
                                 "adj.pval"], 4)
                   ),
           paste0("Adj pval: ",
              round(guild.DFheight.results["grainivore.esr", "adj.pval"], 3)),
           "",
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["ground.for.esr", 
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["ground.for.esr",
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["ground.for.esr",
                                 "adj.pval"], 4)
                   ),
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["treshrb.for.esr",
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["treshrb.for.esr", 
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["treshrb.for.esr",
                                 "adj.pval"], 4)
                   ),
           "",
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["conifer.esr", 
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["conifer.esr", 
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["conifer.esr",
                                 "adj.pval"], 4)
                   ),
           paste0("Adj pval: ",
              round(guild.DFheight.results["mixed.esr", "adj.pval"], 3)),
           paste(
                   "Correlation: ",
                   round(guild.DFheight.results["open.esr", 
                                 "corr.SP"], 3),
                   "Variation Expld: ",
                   round(guild.DFheight.results["open.esr", 
                                 "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.DFheight.results["open.esr",
                                 "adj.pval"], 4))
                   
           ),
  `Pct Imperv on Site` = c("",
                 paste(
                   "Correlation: ",
                   round(guild.imperv.results["omnivore.esr", 
                                "corr.SP"], 2),
                   "Variation Expld: ",
                   round(guild.imperv.results["omnivore.esr", 
                                "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.imperv.results["omnivore.esr",
                                "adj.pval"], 3)
                   ), 
                 paste(
                   "Correlation: ",
                   round(guild.imperv.results["insectivore.esr", 
                                "corr.SP"], 2),
                   "Variation Expld: ",
                   round(guild.imperv.results["insectivore.esr", 
                                "var.explnd"], 3),
                   "Adjusted pval: ",
                   round(guild.imperv.results["insectivore.esr",
                                "adj.pval"], 3)
                   ), 
           paste0("Adj pval: ",
              round(guild.imperv.results["grainivore.esr",
                            "adj.pval"], 3)),
           "",
           paste0("Adj pval: ",
              round(guild.imperv.results["treshrb.for.esr",
                            "adj.pval"], 3)),
           paste0("Adj pval: ",
              round(guild.imperv.results["ground.for.esr",
                            "adj.pval"], 3)),
           "",
           paste(
                   "Correlation:",
                   round(guild.imperv.results["conifer.esr", 
                                "corr.SP"], 2),
                   "Variation Expld:",
                   round(guild.imperv.results["conifer.esr", 
                                "var.explnd"], 3),
                   "Adj pval:",
                   round(guild.imperv.results["conifer.esr",
                                "adj.pval"], 3)
                   ), 
           paste0("Adj pval:",
              round(guild.imperv.results["mixed.esr",
                            "adj.pval"], 3)),
           paste0("Adj pval:",
              round(guild.imperv.results["open.esr",
                            "adj.pval"], 3))
           )
  
)


knitr::kable(pretty.bird.guild.esr,
             "latex",
             booktabs = T,
             caption = "Percent impervious surface on site, presence of stands predating development, and median Douglas-fir height (m) explained variation in effective species richness of bird guilds (diet, foraging location, and forest preference).") %>%
      kable_styling(full_width = F,
           latex_options = c("striped"),
           font_size = 10) %>%
      column_spec(1, bold = T, width = "13em") %>%
      column_spec(2:4, bold = F, width = "11em") %>%
      row_spec(0, bold = T) %>%
      landscape()



```


### Explaining variation in bird community structure

The proportion of overcast site visits was also a significant detection variable explaining variation in both the community composition of all birds observed and of foraging birds observed. After controlling for this, all significant variables for both communities represented the outcome of development, landscaping, and management actions. These variables describe  the native tree community on site (median Douglas-fir height as a proxy for tree age, the presence of stands predating development, and native conifer density, tree community cluster), the native shrub community on site (native shrub effective species richness and for foraging birds only, native shrub density), and percent impervious surface and dead wood abundance on site (see [Appendix B: Birds](#CH08-birdappendix)). None of the neighborhood- or parcel-scale socio-economic variables or neighborhood land cover variables were significant, nor were most variables describing maintenance actions (e.g. herbicide application). 

The best supported model for all bird and foraging bird community composition, as with univariate models, included the proportion of visits overcast and median dominant Douglas-fir height. Multiple other models received equal support, including single variable models with percent impervious surface on site and native conifer density, and multiple two variable models. Other two and three variable models along with all four+ variable models received substantially less support (Table \@ref(tab:CH3-incidence-AICc) and Table \@ref(tab:CH3-foraging-AICc)). The best supported models explain approximately 37-43% of variation. Median dominant Douglas-fir height and native conifer density were consistently included in models with $\Delta$ AICc $\leq$ 2.

``` {r CH3-incidence-AICc}

colnames(pretty.incidence.AICc)[5] <- "Pseudo-F"

#order(pretty.incidence.AICc$`Delta AICc`, decreasing = FALSE)

knitr::kable(pretty.incidence.AICc[ c( 9, 3, 8, 6, 5, 4, 1, 2, 7, 10),
                                    c(1, 7, 5:6,3)],
       booktabs = TRUE,
       caption = "Comparison of bird community PERMANOVA models using AICc.  Models with socio-economic variables were not significant; median household income is presented here for comparison only.") %>%
      kable_styling(full_width = F,
           latex_options = c("striped"),
           font_size = 10) %>%
      column_spec(1, bold = F, width = "16em")

```


``` {r CH3-foraging-AICc}

#pretty.foraging.AICc <- pretty.foraging.AICc[-16, ]
#order(pretty.foraging.AICc$`Delta AICc`, decreasing = FALSE)

colnames(pretty.foraging.AICc)[5] <- "Pseudo-F"


knitr::kable(pretty.foraging.AICc[ c(8,3,6,2,4,1,5,7,9),
                  c(1,7,5:6,3)],
       booktabs = TRUE,
       caption = "Comparison of foraging bird community PERMANOVA models using AICc.  Models with socio-economic variables were not significant; median household income is presented here for comparison only.") %>%
      kable_styling(full_width = F,
           latex_options = c("striped"),
           font_size = 10) %>%
      column_spec(1, bold = F, width = "16em")



```



### Vegetation association preference

Bird species associated with conifer forests, including Brown Creeper and Townsend's Warbler, prefer sites with native tree community clusters [@bnao]. The only bird species associated ornamental tree communities was the American Crow. The habitat preferences of birds observed foraging on site also follows this pattern (Table \@ref(tab:CH3-point-biserial)). 



## Discussion {#CH3-discussion}

Development and landscaping actions alter vegetation communities in social-ecological systems [@faeth2011; @avolio2018; @dysonveg2019]. I find that these outcomes of development and landscaping activities influence the presence and foraging activity of birds on office developments in Bellevue and Redmond, WA. 

### Vegetation association in urban office developments follows other habitats

The bird species with significant ecological preferences in point biserial correlation analysis support my hypothesis that urban vegetation associations and habitat use is consistent with those observed elsewhere [@bnao]. Brown Creeper and Red-breasted Nuthatch are bark foragers reliant on large conifer trees which support their prey species and provide both nesting and wintering habitat. Townsend's Warbler, Golden-crowned Kinglet, and Chestnut-backed Chickadee are canopy foragers primarily in conifer forests. Brown Creepers in particular are a species of concern due to Douglas-fir harvesting, though other species are also under pressure from land use conversion [@bnao]. Diets of Red Breasted Nuthatch and Chestnut-backed Chickadee are largely insectivorous but supplement their diets with conifer seeds [@bnao]. Only American Crow, a generalist urban species, was associated with the ornamental tree habitat cluster.

Overall, conifer associated native bird species preferred habitat on office developments characterized by native trees, including conifers. The observed bird community on office developments is similar to the bird communities observed in 'young' (40-75 y.o.) and 'mature' (105-165 y.o.) Douglas-fir stands in western Washington [@manuwal1987; @haveri2000]. These stand ages correspond with trees planted in the late 1970s following development and preserved Douglas-fir trees that established following logging in the early 1900s. 

Demonstrated use and foraging activity provide evidence that office developments provide resources important to individual fitness for winter birds, and is an important first step in establishing habitat quality, habitat selection, and effects on population size [@jones2001]. My results are consistent with other urban bird research, which found that native bird species were disproportionately associated with native vegetation [@burghardt2009; @lerman2011; @belaire2014; @mackenzie2014; @chong2014]. Additional research quantifying demography (population size, change, and age distribution) and factors influencing habitat selection (e.g. predation, food availability) are needed to definitively assess the habitat quality offered on office developments and their contribution to population size. 


### Variation in bird effective species richness and community composition

I observed very high agreement between univariate regression tests using effective species richness and multivariate tests using community composition. Both analyses identified development and landscaping variables that describe the outcome of developer and land owner decisions to preserve the tree community, resulting land cover, and the resulting tree and shrub community as significant in explaining variation after controlling for the proportion of overcast days. The best model for explaining variation in observed effective species richness and community composition included median dominant Douglas-fir height, after controlling for the proportion of overcast days. This result supports my hypothesis that variables describing the impact of development and landscaping actions on vegetation explain variation in the bird community. However, it does not support my hypothesis that univariate and multivariate analyses would reveal different patterns.




#### Response of feeding guilds {-}

Variation in bird guild effective species richness was variously explained by the three development and landscaping outcomes. Consistent with expectations, conifer associated species were negatively correlated with impervious surface on site and positively correlated with the presence of older, larger trees. This agrees with prior knowledge of birds' habitat preferences and the vegetation preference results of the point-biserial correlation coefficient analysis [@bnao]. For example, Red Crossbill is a mature conifer associated species that relies on conifer seeds to overwinter; those I observed belonged to the Pseudotsuga subgroup and were only observed foraging on Douglas-fir cones. Similarly, insectivorous bird effective species richness was negatively correlated with impervious surface on site and positively correlated with the presence of older, larger trees, consistent with other studies  [@blair2008; @zhou2012].

However, omnivorous birds also followed this pattern, which differs from other studies where birds classified as omnivores in other studies are generally urban adapted species [@chace2006; @barth2015; @clucas2015; @marzluff2016]. In addition to urban adapted species like American Crow and European Starling, omnivorous species in this study include conifer-associated natives like Red-breasted Nuthatch and Chestnut-backed Chickadee. Red-breasted Nuthatch expands its diet of arboreal arthropods to include large number of conifer seeds [@bnao]. Importantly, bird guild assignment varies based on season and location; many species have a more inclusive diet during the winter than during breeding season. For example, Audubon's Warbler is insectivorous during the breeding season, but in the winter expands its diet to include significant amounts of fruit and is thus omnivorous [@bnao]. The same bird species therefore requires different winter and summer resources, and winter and summer habitat cues are likely different [@cody1985]. 

#### Bottom-up development and landscaping variables are important {-}

Both univariate and multivariate models suggest that "bottom-up" patterns resulting from parcel-level decision making are more important than "top-down" patterns that emerge at the neighborhood-scale from diffuse decisions in determining bird community on office developments [@kinzig2005] confirming my hypothesis. Site-scale variables including plant selection also explained more variation than neighborhood-scale variables elsewhere in the Pacific Northwest [@galitsky2015] and on residential properties [@belaire2014; @daniels2006]. These site-scale variables also influence breeding bird diversity and abundance and impact foraging preference [@burghardt2009; @narango2017]. Differences in insect abundance between native and non-native shrubs are frequently identified as a causal factor [@burghardt2009; @tallamy2004; @burghardt2013]. These results suggest that for bird habitat use, social-ecological systems are heterogeneous across organizational units in such a way that small scale interactions between social and ecological spheres are more important than more diffuse patterns, at least for highly mobile species like birds. 


#### Poor support for other bottom-up variables {-}

However, bottom-up socio-economic patterns included in this analysis were not significant in explaining community variation. Most notably, parcel size was not significant here, though area consistently explained biodiversity in residential land uses and parks [@schutz2015; @beninde2015; @evans2009]. Since the species-area relationship is well established, different results may be based on different methodological choices or the relative paucity of bird species native to the Seattle area. This study was designed so that there is heterogeneity of vegetation between study units, but each study unit is relatively homogeneous. When area is important, the study unit may be internally heterogeneous, such that additional habitat types and vegetative complexity are included as area increases [@marzluff2016; @schutz2015]. 

Bottom-up patterns in maintenance inputs were also not significant in explaining measures of bird community. In other studies, irrigation was shown to increases insect prey availability and may increase the diversity of bird communities in arid areas [@schleder2010]. Here, irrigation is very common, reducing my ability to detect differences; bird foraging in the wet winter months may also be unaffected. More research on the impact of maintenance activities on bird populations and foraging activity is needed in urban areas, as these activities in agricultural areas have long been known to impact bird mortality and foraging success [@mineau2013; @lepczyk2004; @henny1972; @fry1995; @mineau1994; @mineau2002; @gibbons2015]. While pesticide and herbicide use on site did not significantly explain variation in measures of the bird community, there may be a critical exposure issues for sites with older native conifers an high bird foraging use that apply these chemicals. Particularly impacted birds would be ground and shrub foragers and insectivores associated with conifer habitat including Pacific Wren and Varied Thrush, along with sparrows and other wrens. Less impacted would be high conifer foragers if most of their prey (insect or plant) is not exposed.

#### Poor support for top-down variables {-}

Top-down neighborhood-scale socio-economic and land cover variables resulting from diffuse decisions made in socio-ecological systems did not receive support in either univariate or multivariate analyses. This result does not support my hypothesis that socio-economic variables significant in other research would be significant here. In studies of residential property, socio-economic neighborhood-scale variables significantly explained variation in bird species richness in studies of residential property [@lerman2011; @luck2013; @melles2003]. Land cover variables, including surrounding vegetation, tree canopy, and impervious surface cover were also significant [@schutz2015; @beumer2015; @vanheezik2013; @melles2003]. However, there is disagreement in the existing literature, with some studies agreeing with my results that neighborhood-scale variables are less or not important for passerine community composition [@belaire2014; @galitsky2015; @daniels2006].

Birds are highly mobile species with ranges that overlap multiple land use types. In Bellevue and Redmond, WA, commercial property is adjacent to residential land use. A bird observed on a commercial parcel also includes residential parcels in its home range. Decisions in social-ecological systems are driven by diverse agents, and decisions on commercial properties are likely driven by different motivations and preferences than residential property. For example, landscaping for self and neighborhood status on residential property versus landscaping to attract tenants on commercial property [@nassauer2009]. The outcome of different decision pathways should theoretically be important to birds only as far as they influence habitat quality. If they impact habitat quality differently, the proportion of different land uses in the area may influence bird communities. 

My results suggest that neighboring residential property habitat either doesn't matter, or is of sufficient quality to not affect passerine birds observed on office developments. It may also be that for local species habitat selection processes, the patch of habitat on the office development is more important than what surrounds it. Further behavioral ecology research and studies examining bird use of adjacent land uses is needed.

Methodological differences may partly explain the difference. Most other studies address other land uses, use neighborhoods instead of management units (parcels) as the unit of analysis, or use other bird count methods. My research also did not examine the ends of the socio-economic gradient, while others do not study the entire vegetation gradient. However, these alone can't explain differences in results, my study did agree with some studies on residential land uses using more standard methods [e.g. @belaire2014]. Standardized study designs across multiple urban areas and multiple land use types are needed to determine which factors are important in determining whether site-scale variables, neighborhood-scale variables, or both are significant. Potential explanatory variables include local bird ecology and life history, breeding vs. non-breeding season [as in @clergeau1998], patterns of development, or some combination of these. 


``` {r CH3-lit-comparison, include = F}

# My results also substantially agree with other studies that have found various measures of forest or canopy cover (including native tree species richness, canopy cover, and vegetation cover) and impervious surface cover significant in explaining variation in bird species richness and community composition 

  lit.compare <- tibble(Author = c("This study", "This study", "@daniels2006", "@schutz2015",
                   "Villegas et al 2010", "Lerman et al 2011", "@germaine1998", "@clergeau1998",
                   "@clergeau1998", "@schutz2015", "@loss2009", "@luck2013", "@luck2013",
                   "@belaire2014", "@belaire2014"),
          `Study unit` = c("Office development only", "250m, 500m, 1km buffer of office development",
             "Residential gardens", "City park", "One hectare area around point count center",
             "Three 100 x 40m transects near point count", "100m buffer around point count",
             "Plots on urbanization gradient", "Plots on urbanization gradient",
             "500m buffer of park centroid", "Various; see study", "1km radius of neighborhood boundary",
             "Within neighborhood", "Within neighborhood", "1km radius of neighborhood boundary"),
          `Approximate area` = c(".5-5 acres", "52-1300 acres", ".01-.4 acres", "1- 85 acres", "2.5 acres",
                      "3 acres", "7.8 acres", "25-50 acres", "25-50 acres", "195 acres",
                      "small plot mean size 1450 acres; large plot mean size 11000 acres",
                      "Various; see study", "Various; see study", "approx 12.4 acres",
                      "approx 1300 acres"),
          Season = c("Winter", "Winter", "Winter and breeding", "Winter", "Breeding",
                "Winter and breeding", "Breeding", "Breeding", "Winter", "Winter", 'Breeding',
                "Spring/summer, some sites year-round", "Spring/summer, some sites year-round",
                "Breeding", "Breeding"),
          Location = c("Seattle, WA, USA", "Seattle, WA, USA", "Hobart, Tasmania", "Vienna, Austria",
                 "La Paz, Bolivia", "Phoenix, AZ, USA", "Tucson, AZ, USA", "Quebec and France",
                 "Quebec and France", "Vienna, Austria", "Chicago, IL, USA",
                 "South-eastern Australia", "South-eastern Australia", "Chicago, IL, USA",
                 "Chicago, IL, USA"),
          Impervious = c("Dec. BSR", "No relationship", "", "", "Dec. BSR", "", 
                  "Inc. Non-native R; Dec. Native R", "", "", "Dec. BSR ",
                  "No relationship with total BSR; Dec. Native R; Dec. Migratory R",
                  "No relationship", "", "", ""),
          `Forest/canopy` = c("Inc. BSR with native conifers", "No relationship", "Inc. BSR with canopy height, decid cover",
                   "Inc. BSR", "Inc. BSR", "Inc. Native richness with native trees", "Dec. Non-native R; Inc. Native R",
                   "Inc. BSR", "No relationship", "", "", "", 
                   "Inc. BSR & native R w/ native nectar plants", "Inc. native BSR w/ tree presence",
                   "No relationship")
    )

knitr::kable(lit.compare,
             caption = 'Comparison of site- and neighborhood-scale percent impervious surface and canopy cover between multiple studies. Inc.: Increase in metric observed; Dec.: Decrease in metric observed; R: R: Richness; BSR: Bird species richness.',
             "latex") %>%
      kable_styling(full_width = F, 
                    latex_options = c("striped"),
                    font_size = 10) %>%
      column_spec(2, width = "11em") %>%
      column_spec(3:5, width = "8em") %>%
      column_spec(6:7, width = "14em") %>%
      landscape()

#, fig.cap = 'Comparison of site and neighborhood scale percent impervious surface cover and canopy cover between multiple studies.'
#
# To make recommendations in urban landscape management that are robust to year-round variation in guild assignment and better support species population ecology, further research is needed to observe how species vegetation associations and requirements shift throughout the year. 

```



### Study strengths and limitations

While differences in methodology makes comparison between this research and other studies more difficult, they contributed to the strengths of this study. First, sampling across the vegetation gradient within one land use allowed for detection of significant differences in tree and shrub communities and for detecting patterns of bird habitat association and important parcel-scale variables. These provide the foundation for future policy and management recommendations. Second, using the standardized search method facilitated research in noisy urban areas when birds were not calling, and kept the unit of analysis as the parcel instead of the area of a point count [@watson2003]. This allowed me to connect parcel-specific management actions with the bird community's habitat use.

There are also important limitations of this study. First, bird studies were conducted during the winter and not year round. Policy and management best practices suggestions made based on winter observations may not capture the stricter summer requirements, and more robust year round research is needed. Second, my site selection did not sample across the full range of surrounding impervious surface and parcel area. This may lead to under-valuing the importance of these variables in models of bird richness and community composition. Finally, habitat use research on its own cannot quantify habitat quality or birds' process of habitat selection without behavioral or life-history information [@sherry1985; @jones2001]. 


### Policy and management recommendations

Overall, my results suggest that bird association with a specific office development is best explained by the vegetation community structure and impervious surface cover on that specific parcel. This does not mean that policy implemented at larger scales is not important, instead my results suggest that a critical role of policy is to encourage parcel-scale decision-making and actions that support birds.  Where residents are interested in conservation, policy should alter developer and landowner actions to modify the resulting vegetation communities on site to better support local bird species populations.

My recommendations are based on patterns observed in this research on winter bird communities and in studies of breeding birds [@belaire2014; @kang2015; @shoffner2018]. Additionally, known bird habitat preference information should be used to manage for habitat used by species of interest [@bnao]. For example, if we are interested in native bird species including Brown Creeper and Townsend's Warbler, we should actively promote the native conifer habitat these species are associated with.

One practical implication of bird's relaxed winter foraging requirements is that a wider range of urban properties may be valuable in the winter. Fulfilling winter resource requirements is a crucial part of bird conservation as overwintering survival and body condition influences breeding success [@faaborg2010; @woodworth2017; @norris2004; @calvert2009].

__Tree retention:__
In the Puget Trough, human actions supporting native conifers are important for the continued presence of native forest species including Brown Creeper, Varied Thrush, and Pacific Wren. Established trees provide critical opportunities for foraging, nesting, and perching, and are more beneficial than newly planted replacement trees [@barth2015; @sung2012]. Minimizing the loss of conifer trees to the maximum extent practicable is supported both by this work and others [@barth2015; @stagoll2010; @shryock2017; @shoffner2018]. Key exceptions are if a site is extensively infected with a root rot (e.g. _Phellinus sulphurascens_, _Heterobasidion_ sp.) or if existing trees are otherwise compromised.

Ideally, city policy would encourage development that avoids "green land" disturbance, and designs that preserve as many trees and as much native soil as possible while also fulfilling client needs. Other researchers' priorities mirror these; they suggest minimizing forest loss first, then reducing disturbance effects in the matrix [@shoffner2018]. Existing tree protection policies are limited by local politics and developer and landowner compliance, and the efficacy of existing policy is mixed [@sung2012; @hill2009]. Other city policies, including parking space requirements and parking lot planting requirements, result in development patterns that favor small cultivated trees over native conifer retention and planting. These policies will need to be changed to better support bird use of sites.

__Planting trees:__
Additionally, actions and policies that accomplish long-term urban forest management---particularly planning for succession---will be critical to the future of urban birds in the Puget Trough and elsewhere [@threlfall2016]. In addition to aligning policy to support native tree planting, outreach and initiatives encouraging larger native tree planting may be necessary to overcome the observed bias towards planting smaller ornamental deciduous trees [@tenneson2014]. While some ornamental non-native trees like seed-bearing _Fraxinus_ sp. may selectively provide important food resources, the inclusion of native conifer variables in well supported models and the results of other research studies suggests that native trees are more important [@schlaepfer2011; @narango2017]. Far more native bird species are associated with the native tree habitat community typology than the ornamental tree habitat community typology, suggesting that planting natives will not negatively impact local bird species, but if these actions are not taken, species could be materially hurt. 

While very young trees provide little habitat value [@sung2012], trees should become more valuable over time as they mature. For example, older neighborhoods with more mature cultivated vegetation supported higher bird diversity than newly developed neighborhoods [@white2005streets], and birds prefer foraging in native trees and that these trees support higher prey abundance [@narango2017]. Planting trees also anticipates existing tree death to maintain the future of the urban forest, as planting largely replaces tree dispersion and establishment processes on landscaped properties [@ettinger2017]. 

__Planting shrubs:__
Planting native shrubs is suggested as a straightforward action to provide food resources to native birds as shrubs are easily planted, grow quickly relative to trees, and provide cover and foraging resources [@cerra2016]. Other studies have found that increasing the volume or complexity of understory vegetation increases urban bird species richness [@threlfall2016]. However, landscaping changes on commercial developments (0.5-5 acres) can be expensive; based on anecdotal evidence these changes are outside of the budget for many of the properties in my sample. Additional research on high native shrub richness sites, including post-installation monitoring, is needed specifically in commercial landscapes to evaluate the cost/benefit of shrub community shifts for birds, insects, and other urban wildlife. For example, Varied Thrush often chooses areas with less shrub cover when foraging [@beck2000].

__Other plantings:__
Further, conifer forests are not the only type of habitat historically found in the Puget Trough. Landscaping on office developments imitating grasslands, young forests, or other ecosystems could provide important resources for native bird species and other wildlife while offering design alternatives for developers and landowners when there is no existing forest or retaining trees on site is not feasible. An important question for future research is whether such designs provide resources for desired bird species and other wildlife. Note that this is almost entirely unexplored. During my vegetation classification visits to 400+ properties in Bellevue and Redmond, no property was landscaped in this fashion. Two key limitations likely exist. First, the aesthetics of grasslands are frequently challenging [@hands2002]. Second, current policy may preclude these alternative native habitats, necessitating municipalities to grant waivers or alter landscaping codes before they can be tested.

  

Overall, changing developer and landowner actions and preferences is likely the only way to sustainably alter development patterns, and there are strong arguments for ecologists' involvement in the development process [@niemela1999; @felson2013; @miller2002]. In addition to outreach for developers and landowners on the importance of trees both to potential clients [@kaplan2007; @lottrup2012; @nesbitt2017] and wildlife [this paper; @barth2015; @stagoll2010; @shryock2017], joint research with developers to identify building techniques that are both cost effective and conserve birds are essential. Meaningful sustainability must include a fundamental shift in how sites are developed---moving from a 'tabula rasa' approach where sites are cleared of vegetation and topsoil prior to development to a more place-based and site-specific design and engineering approach.


## Conclusion

Parcel-scale development and landscaping actions, including tree removal and planting species selection, directly and indirectly alter vegetation communities and create variation in the vegetation communities on commercial land use [@dysonveg2019]. I found that conifer-associated bird species exhibited clear associations with office developments where native trees were preserved through development and planted during site landscaping. Measures of bird effective species richness and community composition were significantly explained by variables describing the age and density of native conifers, the abundance of native shrubs, and the amount of impervious surface present at the office development. These results provide support for human-mediated changes in vegetation affecting higher trophic levels on office developments in Redmond and Bellevue, WA [@faeth2011]. 

My results contribute to our understanding of how human actions at the parcel-scale correlate with bird community composition on commercial land uses, and can help guide policy and management best practices. Evidence that developer and landowner vegetation choices impact bird richness and community composition suggest that actions and policies that accomplish native conifer preservation and the long-term management of the urban forest will be critical to the future of native birds that overwinter in the Puget Trough. These site-scale actions together contribute to the habitat provided by the urban matrix, and increasing habitat quality in the urban matrix contributes to breeding bird success and may reduce the probability of local extinction [@rosenzweig2003; @shoffner2018]. There is a significant role for developers, landowners, and policy makers to shape this future. Specific recommendations for the Puget Trough arising from this research include maximizing preservation of existing native canopy cover during development---including designing buildings to preserve trees and soils and avoiding 'green' development---and landscaping with native conifers. 

Future research efforts should center on developing building techniques that are both cost effective and foster bird conservation, as changing developer and landowner actions and preferences is critical to sustainably and meaningfully altering development patterns. Where native conifers cannot be preserved or have already been lost, researchers should explore the value to local bird species of planting native conifer and deciduous trees and of alternative native habitats, including grasslands and shrublands mimicking newly disturbed/early successional habitat. Post-installation monitoring to determine the efficacy of suggested approaches is critical, though often ignored, and can also discover unintended consequences including elevated bird-window strike rates [@hostetler2011; @hager2017]. Finally, birds move across the landscape and have ranges overlapping multiple land use types. More comprehensive research on bird habitat associations and different land uses that relates management units (parcels) to bird presence is needed. More in depth studies of bird behavior and population biology including tracking their movements across cities is needed to better understand how the complex patterns that emerge in social-ecological systems impact bird habitat use.


# Supplemental Information

Supplemental Table \@ref(tab:pretty-descriptive-site) caption: Summary statistics for all 20 office developments where birds were sampled.

``` {r pretty-descriptive-site}

knitr::kable(pretty.descriptive.birds.site,
             booktabs = TRUE,
             row.names = FALSE, 
             caption = "Summary statistics for all 20 office developments where birds were sampled.")

```

Supplemental Table \@ref(tab:pretty-descriptive-species-table) lists all bird species observed.

``` {r pretty-descriptive-species-table}

    pretty.descriptive.birds.species <- order(pretty.descriptive.birds.species, `Common Name`)

knitr::kable(pretty.descriptive.birds.species[ , -c(4:5,8:9)],
             booktabs = TRUE,
             row.names = FALSE, 
             caption = "All observed bird species.") %>%
kable_styling(full_width = F,
              latex_options = c("striped"),
              font_size = 9) %>%
    column_spec(3:6, width = "5em")


```


All univariate PERMANOVA model results for bird effective species richness. Detection variables, including variables median air temperature through total precipitation in Table \@ref(tab:univariate-model-results) were tested in simple models. All other variables were tested with the % of overcast visits as the control variable.

``` {r univariate-model-results, fig.pos="H"}


#pretty.univ.results <- pretty.univ.results[, c(4, 1, 2, 3, 5)]
# pretty.univ.results <- pretty.univ.results[order(pretty.univ.results$`Delta AICc`, decreasing = FALSE) , ]

pretty.univ.results$pretty.names <- str_remove_all(pretty.univ.results$pretty.names, "__")

rownames(pretty.univ.results) <- pretty.univ.results$pretty.names

colnames(pretty.univ.results)[2:3] <- c("Pseudo-F", "p-value")



pretty.univ.results[, c(1:4,6)] <-
        lapply(pretty.univ.results[, c(1:4,6)], as.numeric) %>%
        lapply(., round, digits = 3)

knitr::kable(pretty.univ.results[, c(1:3,6)],
             booktabs = TRUE,
             "latex",
             caption = "Univariate model results for bird effective species richness. Detection variables are tested alone; the percentage of overcast visits is used as a control variable.") %>%
    kable_styling(full_width =F,
              latex_options = c("striped"),
              font_size = 10) %>%
    column_spec(2:6, width = "5em")

```

